Types from data: \textnormal{Making structured data first-class citizens in F\#}

### 2016-02-01 pldi-2016/paper.tex
Deanonymized version
Adding lub function
Working on final edits

\documentclass[10pt,preprint,blind,clearpagebib]{sigplanconf}

@@

For example, \url{http://openweathermap.org/current} contains one eample to document an end-point 
to get the current weather. Using standard libraries, we might call it as:

\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://api.owm.org/?q=NYC}) \\
 \kvd{match}~\ident{JsonValue.Parse}(\ident{doc})~\kvd{with} \\
 |~\ident{Record}(\ident{root})\rightarrow \\
 \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\
 \quad |~\ident{Record}(\ident{main})\rightarrow \\
 \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\
 \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Lovely \%f!}~\ident{num} \\
 \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 |~\_\rightarrow \ident{failwith}~\str{Incorrect format} 
\end{array}
\end{equation*}
%
The code assumes that the response has a particular shape described in the documentation. The
root node must be a record with a \strf{main} field, which has to be another record containing
a numerical \strf{temp} field. When the shape is different, the code simply fails with an exception. 
While not immediately unsound, the code is manifestly prone to errors if strings are misspelled 
or incorrect shape assumed.

@@

\subsection{Summary}
\label{sec:providers-sum}
Throughout the introduction, we used data sets that demonstrate the typical problems that are frequent
in the real-world (missing data, inconsistent encoding of primitive values and heterogeneous shapes). 
The following JSON response with government debt information returned by the World Bank\footnote{Available at 
\url{http://data.worldbank.org}} demonstrates all three problems:
%
{\small{
\begin{verbatim}
  [ { "page": 1, "pages": 5 },
    [ { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2012", "value": null },
      { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2010", "value": "35.14229" } ] ]
\end{verbatim}
}}
%
\noindent
First of all, the top-level element is a collection containing two values of different shape.
The first is a record with meta-data about the current page and the second is an array with data. 
The actual F\# Data implementation supports a concept of heterogeneous collections (briefly outlined
in \S\ref{sec:impl-hetero}) and provides a type with properties \ident{Record} for the former and 
\ident{Array} for the latter. Second, the the field \ident{value} is \kvd{null} for some records. Third, 
numbers can be represented in JSON as numeric literals (without quotes), but here, they are 
returned as string literals instead.\footnote{This is often used to avoid non-standard numerical 
types of JavaScript.}

@@

\vspace{-0.5em}
\begin{equation*}
% option types
\inference[(nullable)\;]
  {\hat{\sigma_1} \tsep \sigma_2 \vdash \sigma}
  {\kvd{nullable}\langl\hat{\sigma_1}\rangl \tsep \sigma_2 \vdash \kvd{nullable}\langl\dropopt{\sigma}\rangl}
\qquad
\inference[(any)\;]
  { \tytagof(\sigma_1) \neq \tytagof(\sigma_2) & \tytagof(\sigma_1) \neq \ident{nullable} \neq \tytagof{\sigma_2}}
  { \sigma_1 \tsep \sigma_2 \vdash \kvd{any} }  
\end{equation*}
\vspace{-2em}

\begin{equation*}
\hspace{4em}
\inference[(record)\;]
  { \forall i\in\{ 1 .. n \}.(\sigma_i \tsep \sigma'_i \vdash \sigma''_i) }
  { \begin{array}{l}
    \nu \; \{ \nu_1 \!:\! \sigma_1,  \ldots \;, \nu_n \!:\! \sigma_n \} \tsep
    \nu \; \{ \nu'_1 \!:\! \sigma'_1, \; \ldots \;, \nu'_m \!:\! \sigma'_m \} \vdash
    \nu \; \{ \nu_1 \!:\! \sigma''_1, \; \ldots \; , \nu_k \!:\! \sigma''_n \}
    \end{array} }
\end{equation*}
\vspace{-2em}

% all rules are symmetric and reflexive
\begin{equation*}
\hspace{2em}
\inference[(list)\;]
  {\sigma_1 \tsep \sigma_2 \vdash \sigma}
  {[\sigma_1] \tsep [\sigma_2] \vdash [\sigma]}
\quad
\inference[(sym)\;]
  {\sigma_1 \tsep \sigma_2 \vdash \sigma}
  {\sigma_2 \tsep \sigma_1 \vdash \sigma}
%
\quad
\begin{array}{l}
 \textnormal{\footnotesize{(refl)}}\;\; \sigma \tsep \sigma \vdash \sigma\\[0.6em]
 % bottom type
 \textnormal{\footnotesize{(bot)}}\;\; \bot \tsep \sigma \vdash \sigma
\end{array}
%
\quad
% null and primitives
\begin{array}{l}
 \textnormal{\footnotesize{(null)}}\;\; \sigma \tsep \kvd{null} \vdash \addopt{\sigma} \quad (\sigma \neq \bot)\\[0.6em]
 \textnormal{\footnotesize{(prim)}}\;\; \kvd{int} \tsep \kvd{float} \vdash \kvd{float}
\end{array}
\end{equation*}

@@

\begin{figure}
\noindent
\begin{equation*}
\begin{array}{rl}
 \textnormal{\footnotesize{(member)}}&
 \hspace{-0.6em}
 \inference
 { \kvd{type}~C(\overline{x:\tau})= \kvd{member}~N_i : \tau_i = e_i \ldots \in L }
 { L, (\kvd{new}~C(\overline{v})).N_i \reduce e_i[\overline{x} \leftarrow \overline{v}] }\\
 \\[-0.2em]
 \textnormal{\footnotesize{(match1)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~\ident{None}~\kvd{with} \\
  \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_2 \\
 \\[-0.2em]
 \textnormal{\footnotesize{(match2)}}&
 \hspace{-1em}
 \begin{array}{l}
    \kvd{match}~\ident{Some}(v)~\kvd{with} \\
    \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_1[x\leftarrow v]\\
 \\[-0.2em]
 \textnormal{\footnotesize{(match3)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~\kvd{nil}~\kvd{with} \\[0em]
  x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_2\\
 \\[-0.2em]
 \textnormal{\footnotesize{(match4)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~v_1::v_2~\kvd{with} \\[0em]
  x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_1[\overline{x}\leftarrow\overline{v}] \\
 \\[-0.2em]
 \textnormal{\footnotesize{(cond1)}}&
 \hspace{-0.4em}
 \kvd{if}~\kvd{true}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_1 \\
 \\[-0.2em]
 \textnormal{\footnotesize{(cond2)}}&
 \hspace{-0.4em}
 \kvd{if}~\kvd{false}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_2 \\
 \\[-0.3em]
 \textnormal{\footnotesize{(eq1)}}&
 v=v\reduce\kvd{true} \qquad \textnormal{\footnotesize{(eq2)}}~~v=v'\reduce\kvd{false}\\
 \\[-0.3em]
 \textnormal{\footnotesize{(fun)}}&
 \hspace{-0.4em}
 (\lambda x.e)~v ~\reduce~ e[x\leftarrow v] \\
 \\[-0.3em]
 \textnormal{\footnotesize{(ctx)}}&
 \hspace{-0.4em}
  E[e] \reduce E[e'] \qquad\qquad(\textnormal{when}~e \reduce e')\\
\end{array}
\end{equation*}

@@

\paragraph{Type provider mapping.}
A type provider produces an F\# type $\tau$ together with a Foo expression and a collection of 
class definitions. We express it using the following mapping:

\noindent
\begin{equation*}
\sem{-} : \sigma \rightarrow (\tau \times e \times L) \qquad (\textnormal{where}~L,\emptyset \vdash e:\ident{Data}\rightarrow\tau)
\end{equation*}
%
The mapping $\sem{\sigma}$ takes an inferred shape $\sigma$. It returns an F\# type $\tau$ and
a function that turns the input data (value of type \ident{Data}) into a Foo value of type $\tau$. 
The type provider also generates class definitions that may be used by $e$.

@@

\noindent
The progress lemma states that evaluation of a well-typed program does not reach an undefined state. 
This is not a problem for the (standard) ML \cite{sml} subset and object-oriented subset \cite{fwjava} of the calculus. The 
problematic part are the dynamic data operations (Figure~\ref{fig:op-conversions}). Given a data
value (of type \ident{Data}), the reduction can get stuck if the value does not have a structure 
required by a specific operation.

@@

We connect two lines of research: integration of external data into statically-typed 
programming languages and inferring types for real-world data sources. We build on F\# type 
providers \cite{fsharp-typeprov,fsharp-typeprov-ddfp,idris-tp,liteq}, but our paper is novel 
in that it discusses the theory and safety of a concrete type provider.

>>>>

\documentclass[10pt,preprint,clearpagebib]{sigplanconf}

@@

For example, \url{http://openweathermap.org/current} contains one example to document an end-point 
to get the current weather. Using standard libraries, we might call it as\footnote{We use an abbreviation
\url{http://api.owm.org/?q=X} for the full URL:
\url{http://api.openweathermap.org/data/2.5/weather?q=X} 
}:

\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://api.owm.org/?q=NYC"}) \\
 \kvd{match}~\ident{JsonValue.Parse}(\ident{doc})~\kvd{with} \\
 |~\ident{Record}(\ident{root})\rightarrow \\
 \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\
 \quad |~\ident{Record}(\ident{main})\rightarrow \\
 \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\
 \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Lovely \%f!}~\ident{num} \\
 \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
 |~\_\rightarrow \ident{failwith}~\str{Incorrect format} 
\end{array}
\end{equation*}
%
The code assumes that the response has a particular shape described in the documentation. The
root node must be a record with a \strf{main} field, which has to be another record containing
a numerical \strf{temp} field. When the shape is different, the code simply fails with an exception. 
While not immediately unsound, the code is manifestly prone to errors if strings are misspelled 
or incorrect shape assumed.

@@

\subsection{Summary}
\label{sec:providers-sum}
Throughout the introduction, we used data sets that demonstrate the typical problems that are frequent
in the real-world (missing data, inconsistent encoding of primitive values and heterogeneous shapes). 
The following JSON response with government debt information returned by the World Bank\footnote{Available at 
\url{http://data.worldbank.org}} demonstrates all three problems:
%
{\small{
\begin{verbatim}
  [ { "page": 1, "pages": 5 },
    [ { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2012", "value": null },
      { "indicator": "GC.DOD.TOTL.GD.ZS",
        "date": "2010", "value": "35.14229" } ] ]
\end{verbatim}
}}
%
\noindent
First of all, the top-level element is a collection containing two values of different shape.
The first is a record with meta-data about the current page and the second is an array with data. 
The actual F\# Data implementation supports a concept of heterogeneous collections (briefly outlined
in \S\ref{sec:impl-hetero}) and provides a type with properties \ident{Record} for the former and 
\ident{Array} for the latter. Second, the field \ident{value} is \kvd{null} for some records. Third, 
numbers can be represented in JSON as numeric literals (without quotes), but here, they are 
returned as string literals instead.\footnote{This is often used to avoid non-standard numerical 
types of JavaScript.}

@@

{\color{red} TO BE DELETED}

\vspace{-0.5em}
\begin{equation*}
% option types
\color{red}
\inference[(nullable)\;]
  {\hat{\sigma_1} \tsep \sigma_2 \vdash \sigma}
  {\kvd{nullable}\langl\hat{\sigma_1}\rangl \tsep \sigma_2 \vdash \kvd{nullable}\langl\dropopt{\sigma}\rangl}
\qquad
\inference[(any)\;]
  { \tytagof(\sigma_1) \neq \tytagof(\sigma_2) & \tytagof(\sigma_1) \neq \ident{nullable} \neq \tytagof{\sigma_2}}
  { \sigma_1 \tsep \sigma_2 \vdash \kvd{any} }  
\end{equation*}
\vspace{-2em}

\begin{equation*}
\color{red}
\hspace{4em}
\inference[(record)\;]
  { \forall i\in\{ 1 .. n \}.(\sigma_i \tsep \sigma'_i \vdash \sigma''_i) }
  { \begin{array}{l}
    \nu \; \{ \nu_1 \!:\! \sigma_1,  \ldots \;, \nu_n \!:\! \sigma_n \} \tsep
    \nu \; \{ \nu'_1 \!:\! \sigma'_1, \; \ldots \;, \nu'_m \!:\! \sigma'_m \} \vdash
    \nu \; \{ \nu_1 \!:\! \sigma''_1, \; \ldots \; , \nu_k \!:\! \sigma''_n \}
    \end{array} }
\end{equation*}
\vspace{-2em}

% all rules are symmetric and reflexive
\begin{equation*}
\color{red}
\hspace{2em}
\inference[(list)\;]
  {\sigma_1 \tsep \sigma_2 \vdash \sigma}
  {[\sigma_1] \tsep [\sigma_2] \vdash [\sigma]}
\quad
\inference[(sym)\;]
  {\sigma_1 \tsep \sigma_2 \vdash \sigma}
  {\sigma_2 \tsep \sigma_1 \vdash \sigma}
%
\quad
\begin{array}{l}
 \textnormal{\footnotesize{(refl)}}\;\; \sigma \tsep \sigma \vdash \sigma\\[0.6em]
 % bottom type
 \textnormal{\footnotesize{(bot)}}\;\; \bot \tsep \sigma \vdash \sigma
\end{array}
%
\quad
% null and primitives
\begin{array}{l}
 \textnormal{\footnotesize{(null)}}\;\; \sigma \tsep \kvd{null} \vdash \addopt{\sigma} \quad (\sigma \neq \bot)\\[0.6em]
 \textnormal{\footnotesize{(prim)}}\;\; \kvd{int} \tsep \kvd{float} \vdash \kvd{float}
\end{array}
\end{equation*}

\begin{equation*}
\begin{array}{rcl}
\ident{lub}(\sigma, \sigma) &=& \sigma \\
\ident{lub}([\sigma_1], [\sigma_2]) &=& [\ident{lub}(\sigma_1, \sigma_2)]  \\
\ident{lub}(\bot, \sigma) = \ident{lub}(\sigma, \bot) &=& \sigma \\
\ident{lub}(\kvd{null}, \sigma) = \ident{lub}(\sigma, \kvd{null}) &=& \addopt{\sigma} \\
\ident{lub}(\kvd{any}, \sigma) = \ident{lub}(\sigma, \kvd{any}) &=& \kvd{any} \\
\ident{lub}(\kvd{float}, \kvd{int}) = \ident{lub}(\kvd{int}, \kvd{float}) &=& \kvd{float} \\
\ident{lub}(\sigma_2, \kvd{nullable}\langl \hat{\sigma_1} \rangl) = \ident{lub}(\kvd{nullable}\langl \hat{\sigma_1} \rangl, \sigma_2) &=& \addopt{\ident{lub}(\hat{\sigma_1}, \sigma_2)} \\
\ident{lub}(\nu \; \{ \nu_1 \!:\! \sigma_1,  \ldots \;, \nu_n \!:\! \sigma_n \}, 
  \nu \; \{ \nu_1 \!:\! \sigma'_1, \; \ldots \;, \nu_n \!:\! \sigma'_n \}) &=& 
  \nu \; \{ \nu_1 \!:\! \ident{lub}(\sigma_1, \sigma'_1), \; \ldots \; , \nu_n \!:\! \ident{lub}(\sigma_n, \sigma'_n) \}  \\[0.75em]
\textnormal{Otherwise}~~~~ \ident{lub}(\sigma_1, \sigma_2) &=& \kvd{any} \\

\end{array}
\end{equation*}

@@

\begin{figure}
\noindent
\begin{equation*}
\begin{array}{rl}
 \textnormal{\footnotesize{(member)}}&
 \hspace{-0.6em}
 \inference
 { \kvd{type}~C(\overline{x:\tau})= \kvd{member}~N_i : \tau_i = e_i \ldots \in L }
 { L, (\kvd{new}~C(\overline{v})).N_i \reduce e_i[\overline{x} \leftarrow \overline{v}] }\\
 \\[-0.2em]
 \textnormal{\footnotesize{(match1)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~\ident{None}~\kvd{with} \\
  \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_2 \\
 \\[-0.2em]
 \textnormal{\footnotesize{(match2)}}&
 \hspace{-1em}
 \begin{array}{l}
    \kvd{match}~\ident{Some}(v)~\kvd{with} \\
    \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_1[x\leftarrow v]\\
 \\[-0.2em]
 \textnormal{\footnotesize{(match3)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~\kvd{nil}~\kvd{with} \\[0em]
  x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_2\\
 \\[-0.2em]
 \textnormal{\footnotesize{(match4)}}&
 \hspace{-1em}
 \begin{array}{l}
  \kvd{match}~v_1::v_2~\kvd{with} \\[0em]
  x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
 \end{array} \hspace{-0.5em} ~\reduce~ e_1[\overline{x}\leftarrow\overline{v}] \\
 \\[-0.2em]
 \textnormal{\footnotesize{(cond1)}}&
 \hspace{-0.4em}
 \kvd{if}~\kvd{true}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_1 \\
 \\[-0.2em]
 \textnormal{\footnotesize{(cond2)}}&
 \hspace{-0.4em}
 \kvd{if}~\kvd{false}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_2 \\
 \\[-0.3em]
 \textnormal{\footnotesize{(eq1)}}&
 v=v'\reduce\kvd{true} \qquad (\textnormal{when}~v = v')\\
 \\[-0.3em]
 \textnormal{\footnotesize{(eq2)}}&
 v=v'\reduce\kvd{false} \qquad (\textnormal{when}~v \neq v')\\
 \\[-0.3em]
 \textnormal{\footnotesize{(fun)}}&
 \hspace{-0.4em}
 (\lambda x.e)~v ~\reduce~ e[x\leftarrow v] \\
 \\[-0.3em]
 \textnormal{\footnotesize{(ctx)}}&
 \hspace{-0.4em}
  E[e] \reduce E[e'] \qquad\qquad(\textnormal{when}~e \reduce e')\\
\end{array}
\end{equation*}

@@

\paragraph{Type provider mapping.}
A type provider produces an F\# type $\tau$ together with a Foo expression and a collection of 
class definitions. We express it using the following mapping:
%
\begin{equation*}
\sem{\sigma} = (\tau, e, L) \qquad (\textnormal{where}~L,\emptyset \vdash e:\ident{Data}\rightarrow\tau)
\end{equation*}
%
The mapping $\sem{\sigma}$ takes an inferred shape $\sigma$. It returns an F\# type $\tau$ and
a function that turns the input data (value of type \ident{Data}) into a Foo value of type $\tau$. 
The type provider also generates class definitions that may be used by $e$.

@@

\noindent
The progress lemma states that evaluation of a well-typed program does not reach an undefined state. 
This is not a problem for the Standard ML \cite{sml} subset and object-oriented subset \cite{fwjava} of the calculus. The 
problematic part are the dynamic data operations (Figure~\ref{fig:op-conversions}). Given a data
value (of type \ident{Data}), the reduction can get stuck if the value does not have a structure 
required by a specific operation.

@@

We connect two lines of research: the work: inferring types for real-world data sources
and the integration of external data into statically-typed programming langauges.

Early work on F\# Data happened together with the development of the type providers mechanism in F\# 
\cite{fsharp-typeprov,fsharp-typeprov-ddfp}, but previous publications on type providers focused on 
the general mechanism \cite{idris-tp,liteq} rather than on concrete type providers. This paper 
is novel in that it discusses the theory and safety of a concrete type provider from the programming
language perspective.
